---
title: 'Week 1: Solutions'
author: "ETC3580"
output:
  html_document:
    fig_height: 5
    fig_width: 8
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
    number_sections: false
    code_folding: show
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, warning=FALSE, message=FALSE)
options(width=80,digits=3)
```

```{r loadpackages}
library(tidyverse)
library(visreg)
```

# Exercise 1A

## A1

```{r A1}
swiss <- as_tibble(swiss)
```

## A2

```{r A2}

GGally::ggpairs(swiss)
ggplot(gather(swiss)) +
  geom_boxplot(aes(x=key,y=value))
```

 * Fertility vs Agriculture: positively correlated
 * Fertility vs Infant mortality: positively correlated
 * Fertility vs Examination: negatively correlated
 * Fertility vs Education: negatively correlated
 * Fertility vs Catholic: seems to have two groups of provinces

In addition, there are relationships between Agriculture and Education, Agriculture and Examination, and Examination and Education, which are very intuitive.

There are few outliers in Education, Fertility and Infant mortality.

The shape of the boxplot for Catholic is because of having roughly two groups of provinces. You can clearly see it from the density plot of Catholic variable.

## A3

```{r A3}
fit <- lm(Fertility ~ ., data = swiss)
summary(fit)
visreg(fit, "Agriculture", gg=TRUE)
visreg(fit, "Infant.Mortality", gg=TRUE)
visreg(fit, "Examination", gg=TRUE)
visreg(fit, "Education", gg=TRUE)
visreg(fit, "Catholic", gg=TRUE)
```

Agriculture, Education, Catholic, and Infant.Mortality variables look significant.

## A4

```{r A4}
drop1(fit, test="F")
```

Sure enough, only `Examination` is not significant.

## A5

```{r A5}
# Using step
fit2 <- lm(Fertility ~ (.)^2, data=swiss) %>% step()
summary(fit2)
```

These two approaches give almost the same model, with the difference being in how they include the `Catholic` variable.

```{r A5c}
visreg2d(fit2, "Agriculture", "Examination")
visreg2d(fit2, "Agriculture", "Education")
visreg2d(fit2, "Agriculture", "Infant.Mortality")
visreg2d(fit2, "Examination", "Education")
visreg2d(fit2, "Examination", "Infant.Mortality")
visreg2d(fit2, "Catholic", "Education")
```

# Exercise 1B

```{r Bgss}
load("gss.RData")
gss %>%
  # Select variables we want to use
  select(prestige, year, age, sex, race) %>%
  # Convert prestige and age to numeric variables
  mutate(prestige = as.numeric(substr(prestige,1,2)),
         age = as.numeric(substr(age, 1, 2))) ->
  # Save result
  mydata

summary(mydata)

GGally::ggpairs(mydata)
```

## B1

```{r B1}
# Fit model
fit <- lm(prestige ~ year + age + sex + race, data=mydata)
summary(fit)
```

Let's do a standard residual vs fitted plot.

```{r B1a}
fit %>%
  broom::augment() %>%
  ggplot() +
    geom_point(aes(x=.fitted, y=.resid))
```

The two clusters probably occur because of a factor variable with two levels --- mostly likely race or sex. We can check if that is the case.

```{r B1b}
fit %>%
  broom::augment() %>%
  ggplot() +
    geom_point(aes(x=.fitted, y=.resid, color=race))
```

## B2

We will add degree and class. Other choices are also possible, of course.

```{r B2}
gss %>%
  # Select variables we want to use
  select(prestige, year, age, sex, race, degree, class) %>%
  # Convert prestige and age to numeric variables
  # and recode degree and class to fix missing values
  mutate(
    prestige = as.numeric(substr(prestige,1,2)),
    age = as.numeric(substr(age, 1, 2)),
    degree = recode(degree,
                "JUNIOR COLLEGE" = "HIGH SCHOOL",
                "DK" = NA_character_,
                "NA" = NA_character_),
   class = recode(class,
                "IAP" = NA_character_,
                "NO CLASS" = NA_character_,
                "DK" = NA_character_,
                "NA" = NA_character_)) ->
    mydata

fit2 <- lm(prestige ~ year + age + sex + race + degree + class, data=mydata)
summary(fit2)
```

## B3

```{r B3}
fit3 <- lm(prestige ~ year + age * sex + race + degree + class +
             age * sex, data=mydata)
summary(fit3)
```

## B4

We will include all possible interactions, and then use stepwise selection to simplify the model.


```{r B4}
fit4 <- lm(prestige ~ (year + age + sex + race + degree + class)^2,
           data=mydata) %>%
        step(trace=FALSE)
summary(fit4)
```

That's a lot of interactions!

```{r B4b}
visreg2d(fit4, "age", "year")
visreg(fit4, "age", "sex", gg=TRUE, overlay=TRUE)
visreg(fit4, "age", "race", gg=TRUE, overlay=TRUE)
visreg(fit4, "age", "degree", gg=TRUE, overlay=TRUE)
visreg(fit4, "age", "class", gg=TRUE, overlay=TRUE)
visreg(fit4, "year", "sex", gg=TRUE, overlay=TRUE)
visreg(fit4, "year", "race", gg=TRUE, overlay=TRUE)
visreg(fit4, "year", "degree", gg=TRUE, overlay=TRUE)
visreg(fit4, "year", "class", gg=TRUE, overlay=TRUE)
```

Other numeric-categorical interactions can be plotted similarly.

```{r B4c}
visreg(fit4, "sex", "race", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
visreg(fit4, "class", "sex", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
visreg(fit4, "class", "race", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
visreg(fit4, "class", "degree", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
visreg(fit4, "race", "degree", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
visreg(fit4, "sex", "degree", gg=TRUE, points.par=list(alpha=0.1)) +
  theme_bw()
```


## B5

The best model obtained (`fit4`) has AIC `r AIC(fit4)`.

## B6

It has an $R^2$ of 36.71%

## B7

```{r B7}
filter(mydata, year > 2000) %>% summary
```

All `prestige` values are missing. So no data left to model.

## B8

First we need to make sure the comparisons are fair, so we will remove rows with missing values.

```{r B8}
mydata <- na.omit(mydata)
lm(prestige ~ year, data = mydata) %>% AIC()
lm(prestige ~ age, data = mydata) %>% AIC()
lm(prestige ~ sex, data = mydata) %>% AIC()
lm(prestige ~ race, data = mydata) %>% AIC()
lm(prestige ~ degree, data = mydata) %>% AIC()
lm(prestige ~ class, data = mydata) %>% AIC()
```

Degree is the single most important predictor based on AIC.



